Compositional Equivalences Based on Open pNets

### 2022-10-05 2016-Weak-FH-Bisim/WeakBisim.tex
comments

\section{Conclusion and discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of  parallel or distributed systems. One  strength of pNets is their parameterised nature, making them suitable for to the representation of systems of arbitrary size, and making the modelling of parameterised system possible. Parameters are also crucial to reason on interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HenrioKSM:2016}. VCE is the specification and verification platform that uses pNets as an intermediate representation.
\TODO{Eric: short mentions and links to our tools and examples papers}

Open pNets are pNets with holes; they are adapted to represent processes parameterised by the behaviour of other processes, like composition operators or interaction 
protocols that synchronise the actions of processes that can be plugged afterwards. Open
pNets are hierarchical composition of automata with holes and parameters. We
defined here a semantics for open pNets and a complete bisimulation theory for them. 
The semantics of open pNets relies on the definition of open automata that are automata with holes and parameters, but no hierarchy \TODO{Ludo: flatten}. Open automata are somehow labelled transition systems with parameters and holes, a notion that is useful to define semantics, but makes less sense when modelling a system, compared to pNets. To be precise, it is on open automata that we define our FH-bisimilarity relations.

This article defines a strong and a weak bisimulation relation that are adapted to parameterised systems and hierarchical composition. Our FH-bisimulation principle handles pNet parameters in the sense that two states might be or not in relation depending on the value of parameters. Our strong  FH-bisimilarity is compositional by nature in the sense that FH-bisimilarity is maintained when composing processes. We also identified a simple and realistic condition on the semantics of non-observable actions that allows weak FH-bisimilarity to be also compositional \TODO{Eric: recall on rule formats}. Overall we believe that this article paved the way for a solid theoretical foundation for compositional verification of parallel and distributed systems.

pNets support the refinement checking at the automata level through a simulation approach, with symbolic evaluation of
the guards and transitions. 
 The definition of simulation on open automata should be  stronger than a strict simulation since it matches a transition with a family of transitions. Such a relation should be able to check the refinement between two open automata with the same level of abstraction but specified differently, for example, by duplicating states, removing transitions, reinforcing guards, modifying variables. Additionally, composition of 
pNets gives the possibility to either add new holes to a system or  fill holes. A useful simulation relation  should thus support the  comparison of automata that do not have the same number of holes. Designing such a simulation relation is a non-trivial extension of this work that we are investigating.

We are currently extending this work,  looking at further properties of FH-bisimulation, but also
the relations with existing equivalences on both closed and open systems. In particular, our model being significantly different from those considered in \cite{IngolfsdottirL:2001}, it
would be interesting to compare our ``FH'' family of bisimulations with the hierarchy of symbolic bisimulations from these authors.
We also plan to apply open pNets to the study of complex composition
operators in a symbolic way, for example in the area of parallel
skeletons, or distributed algorithms.
We have developed tool support for computing the
symbolic semantics in term of open automata \cite{QBMZ-AVOCS18}, and have
developed algorithms to check strong FH-bisimulation and weak FH-bisimulation \cite{hou:hal-02406098,WangMZ:2021}.
The challenges here, in the context of our symbolic systems, is not so much algorithmic complexity,
as was the case with classical weak bisimulation on finite models, but decidability and termination.
The naive approach using an explicit construction of the weak transition, may in itself introduce non-termination,
so we prefer a direct implementation of the weak bisimulation definition, without
constructing the weak automata, but searching \emph{on demand} to construct the required weak transitions.
Beside, we explore in \cite{wang:hal-03126313} more pragmatic approaches using weak bisimulation preserving (pattern-based) 
reduction rules.

@@

We introduce the following predicate (we will need it for reasoning on the global predicate and will reason on it along the proof):\\
$\begin{array}{l}\displaystyle{
\Pred_\beta=\bigwedge_{p=0}^{m_1}(\beta_{1pj_0}=\alpha_{1p})\subst{\bigodot_{j=p-1}^{0}\!\!\Post_{1j}} \wedge (\beta_{2j_0}=\alpha_{20})\subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}} \wedge \\\qquad\qquad
\displaystyle{
\bigwedge_{p=0}^{m_2}(\beta_{3pj_0}=\alpha_{3p})
\subst{\bigodot_{j=p-1}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}}
\end{array}$

>>>>

\section{Conclusion and discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of  parallel or distributed systems. One  strength of pNets is their parameterised nature, making them suitable for to the representation of systems of arbitrary size, and making the modelling of parameterised systems possible. Parameters are also crucial to reason about interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HenrioKSM:2016}. VCE is the specification and verification platform that uses pNets as an intermediate representation.
\TODO{Eric: short mentions and links to our tools and examples papers}

Open pNets are pNets with holes; they are adapted to represent processes parameterised by the behaviour of other processes, like composition operators or interaction 
protocols that synchronise the actions of processes that can be provided afterwards. Open
pNets are hierarchical composition of automata with holes and parameters. We
defined here a semantics for open pNets and a complete bisimulation theory for them. 
The semantics of open pNets relies on the definition of open automata that are automata with holes and parameters, but no hierarchy. \TODO{check this sentence} Open atomata are a flattened view of the pNet; their behaviour is expressed as open transitions that allow for a more semantic interpretation of parameters than pNets. Open automata are  labelled transition systems with parameters and holes, a notion that is useful to define semantics, but makes less sense when modelling a system, compared to pNets. To be precise, it is on open automata that we define our FH-bisimilarity relations.

This article defines a strong and a weak bisimulation relation that are adapted to parameterised systems and hierarchical composition. Our FH-bisimulation principle handles pNet parameters in the sense that two states might be or not in relation depending on the value of parameters. Our strong  FH-bisimilarity is compositional  in the sense that FH-bisimilarity is maintained when composing processes. We also identified a simple and realistic condition on the semantics of non-observable actions that allows weak FH-bisimilarity to be also compositional \TODO{Eric: recall on rule formats}. Overall we believe that this article paved the way for a solid theoretical foundation for compositional verification of parallel and distributed systems.

pNets support the refinement checking at the automaton level through a simulation approach, with symbolic evaluation of
the guards and transitions. 
 The definition of simulation on open automata should be  stronger than a strict simulation since it matches a transition with a family of transitions. Such a relation should be able to check the refinement between two open automata with the same level of abstraction but specified differently, for example, by duplicating states, removing transitions, strengthening guards, modifying variables. Additionally, composition of 
pNets gives the possibility to either add new holes to a system or  fill holes. A useful simulation relation  should thus support the  comparison of automata that do not have the same number of holes. Designing such a simulation relation is a non-trivial extension of this work that we are investigating.

We are currently extending this work,  looking at further properties of FH-bisimulation, but also
the relations with existing equivalences on both closed and open systems. In particular, our model being significantly different from those considered in \cite{IngolfsdottirL:2001}, it
would be interesting to compare our ``FH'' family of bisimulations with the hierarchy of symbolic bisimulations from those authors.
We also plan to apply open pNets to the study of complex composition
operators in a symbolic way, for example in the area of parallel
skeletons, or distributed algorithms.
We have developed tool support for computing the
symbolic semantics in term of open automata \cite{QBMZ-AVOCS18}, and have
developed algorithms to check strong FH-bisimulation and weak FH-bisimulation \cite{hou:hal-02406098,WangMZ:2021}.
The challenges here, in the context of our symbolic systems, are not so much algorithmic complexity,
as was the case with classical weak bisimulation on finite models, but decidability and termination.
The naive approach using an explicit construction of the weak transition, may in itself introduce non-termination,
so we prefer a direct implementation of the weak bisimulation definition, without
constructing the weak automata beforehand, but searching \emph{on demand} to construct the required weak transitions.
Beside, we explore in \cite{wang:hal-03126313} more pragmatic approaches using weak bisimulation preserving (pattern-based) 
reduction rules.

@@

We introduce the following predicate (we will need it for reasoning about the global predicate and will reason about it along the proof):\\
$\begin{array}{l}\displaystyle{
\Pred_\beta=\bigwedge_{p=0}^{m_1}(\beta_{1pj_0}=\alpha_{1p})\subst{\bigodot_{j=p-1}^{0}\!\!\Post_{1j}} \wedge (\beta_{2j_0}=\alpha_{20})\subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}} \wedge \\\qquad\qquad
\displaystyle{
\bigwedge_{p=0}^{m_2}(\beta_{3pj_0}=\alpha_{3p})
\subst{\bigodot_{j=p-1}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}}
\end{array}$
